TIMES Ireland Model: Residential Sector

### 2021-05-19 manuscript.tex
Updates

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}
\label{Intro:Climate Targerts}
\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
The EU's updated first Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, to met the EU's updated NDC, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.35]{Figures/Non-ETS-Share.png} 
 \caption{Overview of Ireland's 2030 Climate Targets}
 \label{fig:IrelandClimatetargets}
\end{figure}

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per captia at more than twice the average EU27 \cite{EurostatReal}, the highest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and high dependency on fossil fuel imports with little electricity or gas grid interconnections. Under (EU) 2020/2126 - CHECK Ireland has been afforded higher flexibility allowances in non-ETS targets, to compensate for some of these barriers. The Climate Change Advisory Council (CCAC) will advise the government on setting sectoral five year carbon budgets, which comply with all of Ireland's climate targets. The most ambitious targets.... overview in Fig \ref{fig:GHG2030A}, shows some of the most ambitious GHG targets globally and the share of GHG emissions from agriculture. Despite a large share of agriculture emissions, Ireland has set one of the most ambitious targets towards 2030. The CCAC will rely of some modelling tools to provide insights specific to Ireland, one of those tools is discussed in the next section.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.6]{Figures/GHGtargetAGR1.png} 
 \caption{Most Ambitious 2030 GHG Targets}
 \label{fig:GHG2030AGR}
\end{figure}

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
A Energy System Optimisation Model (ESOM) can be a key tool in developing least cost decarbonisation pathways. ESOMs can take account of a large number of dynamic variables, and complex interconnections in each time period and economic sector. The Integrated MARKAL-EFOM System (TIMES) modelling framework is the ESOM developed in this paper. TIMES has been developed and maintained by ETSAP (Energy Technology Systems Analysis Program) under the aegis of the International Energy Agency (IEA), with the key focus to \textit{``establish, maintain and expand a consistent multi-country energy/economy/environment/ engineering analytical capability mainly based on TIMES"} \cite{InternationalEnergyAgency2015InternationalIA}. The contracting parties of ETSAP are the governments of eighteen countries and the European Commission.

\par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as  EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. TIMES is the most understood and used, and because of the expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, TIMES is the preferred ESOM. TIMES is a ``hybrid" ESOM with combines high technological detail and macroeconomic variables covering all energy sectors. The previous Irish TIMES model was extracted from the PET36 model (Pan European TIMES model) and was then updated with national data and assumptions, specific to Ireland. This previous Irish TIMES model was used extensively in the past to provide input to energy and climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Resources2015Irelands2015-2030,DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019,Deane2013TechnicalIreland,IrishGovernment2017National2017},however there was a need to design an ESOM to specifically explore Ireland's long term climate targets and a new model called TIMES Ireland Model (TIM) was built in 2021 to address this need. TIMES has also been used in UK \cite{Li2018IncorporatingModel,Kannan2009ModellingApproaches}, Denmark \cite{Petrovic2016ResidentialSystem,Tattini2018ImprovingModel}, Sweden \cite{Krook-Riekkola2014ModellingStudy,Aberg2011OptimisationBuildings} and China \cite{Shi2016ModellingModel} to find least cost pathways with a focus on the residential sector.While the many climate targets further add to complexity, the need for an ESOM is strengthened. An ESOM can provide insights to least cost energy system pathways, the energy system  transition policymakers must also consider the maximum benefits to the economy to truly optimise the transition. TIM can provide vital input to the discussions or further modelling required, and thus the transparency of an ESOM is critical to ensure absorptive capacity of all stakeholders.

@@

%Preliminary emissions data published by the EPA in January 2021 indicate that emissions from the residential sector increased by 9% in 2020, primarily due to the increase in home working and Covid-related restrictions on movement forcing people to spend more time at homeThese estimates show Residential emissions increased by 9per cent (0.6 Mt CO2eq) in 2020, largely due to the impact of people working from home.  2020 also saw low prices for home heating fuel,
Under the United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting guidelines, set out in ( decision 24/CP.19 ), states the 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Ireland’s residential sector accounted for 14.2\%of all GHGs in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Fig. \ref{fig:Ireland2018GHG} shows an overview of Irelands GHG emissions, with a focus on the energy related GHG emissions. The residential sector accounts for 47\% of heat emissions 30\% of electricity emissions, therefore the residential sector plays a pivotal role in the energy-related GHG emissions in Ireland.

\begin{figure}[h]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside, radius = 2.7]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.8,-0.3) -- (6.8,-.3);
    \hspace{9cm}
    \pie[text=inside,radius=2.2,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/ Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
    \label{fig:Ireland2018GHG}
\end{figure}

 While the national GHG should reduce 51\% by 2030, the residential sector is expected to be reduce GHGs even further, due to the higher cost-effectiveness when compared with other sectors. The high ambition in residential is due to its dependency on receiving low carbon intensity electricity. The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 gCO2/kWh in 2018, 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. However the on-shore wind energy capacity is almost saturated, and policy looks to guide investment into off-shore wind to continue a reduction in carbon intensity. Under CAP2019 this continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but heat and transport also.

@@

\par
The CAP2019 had a Marginal Abatement Cost Curve (MACC) is based on  based on Environmental Protection Agency (EPA) GHG emission projection report \cite{EPA2018EPAReport} and provides a solid analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The MACC explores technological, behavioural and fuel switching. The x-axis (width of each bar chart) shows the potential reduction of annual MtCO2eq. emissions in 2030 from the technology or fuel switch. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of CO2eq. over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in EUR/tCO2eq \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. In the least-cost pathway Ireland would increase Onshore wind by 8.2 GW and Offshore would grow to 3.5 GW,this would be expected to save 7 to 8 MtCO2eq. This would help to indirectly reduce the GHG in the residential sector. However specific residential solutions in the MACC include: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}
\end{itemize}

McKinsey and Company's modelling tools have helped develop the MACC shown in Fig.\ref{fig:CAP2019:MACC}, with more ambitious Non-ETS 2030 target due this year and a national GHG target for 2030, there is a requirement for a new MACC, from which a new residential five-year carbon budgets will arise. 
Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners, while new homes have a strict regulation to comply with, so heat pumps despite their high capital costs are preferred. The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}. The CAP2019 plan to deliever a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.

\subsection{Ireland: Residential Sector Background}
\label{Intro:IreRSDBack}
Among European Union (EU) member states, Ireland has the lowest renewable heat of 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.
\par
Ireland’s decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector, this is largely down to Ireland’s high reliance on fossil fuels and settlement patterns of low thermal efficiency of the housing stock. Ireland's lower than average urbanisation which is connected to only 5.8\% of the population of Ireland living in apartments in 2018, the lowest and well below the average EU-27 of 25.3\% \cite{EurostatDistributionSurveyilc_lvho01}, the lower density relates translates to more expensive mitigation. 33\% of the population of the EU were deemed to live in dwellings too large for their needs in 2018, in Ireland the rate was 71.4\% \cite{EurostatShareSurveyilc_lvho50a}.  
Ireland set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first of strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, is a top priority".}, which will indirectly effect both electricity and transport GHG emission. 
In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved in the last 30 years, and in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018. However, the decline in peat and coal has been offset by more significant growth in oil boilers in rural dwellings in the same period, where oil accounts for 34.8\% of residential GHG emissions \cite{Howley2005ENERGY-RELATEDIreland}. Urban dwellings prefer gas as a heating fuel, which is cheaper, more secure ( 61\% of gas in 2018 was indigenous ) and less carbon intensive. Gas accounted for 15.5\% of residential GHG emission in 2018, significantly lower than oil due to the lower carbon intensity. 
Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}
In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}.

@@

Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock. As mentioned in section \ref{Intro:IreRSD}, Irelands new buildings produce 70\% less GHG emissions than building in 2005, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock highlights the impact the year of construction has on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.The BERs are categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .

@@

\subsection{Current Energy Policy}
\label{Intro:Policy}
The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings
500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent
Installation of 600,000 heat pumps, with 400,000 of these in existing buildings
Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes
Complete the rollout of the Support Scheme for Renewable Heat
Increasing the number of Sustainable Energy Communities to 1,500

@@

\subsection{TIMES Ireland Model (TIM)}
\label{Intro:TIM}
\par
TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets. The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data. The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM . TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing. The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps. TIM will also be open platform to help improve the model and provide transparency. TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy. The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM. for these reasons TIM residential sector is a state of the art modelling platform. 
%GRAPH OF IRELAND DECARBONISATION TARGETS  VS OTHER COUNTRIES - DIFFERENT INTRASTATE and RESOURCE AT BIDENS ENERGY SUBMIT - DENMAKR TOP 2018, 
	ESOM can be used to start the conversation on..
Look at other ESOM, brief explain…
Other residential sectors..
CAP2019 - Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\section{Literature Review}
\label{LitR}
``How should Ireland set GHG targets in the residential sector?” 
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject}

%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

@@

\section{Sample Appendix Section}
\label{sec:sample:appendix}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor section \ref{sec:Introduction} incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

>>>>

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}
\label{Intro:Climate Targerts}
\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
The EU's updated Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.4]{Figures/Non-ETS-Share.png} 
 \caption{Overview of Ireland's 2030 Climate Targets}
 \label{fig:IrelandClimatetargets}
\end{figure}

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports. Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for some of these barriers. To put Irelands targets in perspective the ``Leaders' Summit on Climate" in April 2021, brought together the largest GHG emitters to update and raise ambition of their 2030 climate targets. Comparing these new targets with the same base year of 2018, Fig \ref{fig:GHG2030AGR} shows Ireland's national climate target is ambitious, especially when the share of agriculture GHG emissions is considered. Ireland has some distinct barriers and ambitious targets, the Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets, one of those modelling tools is discussed in the next section.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.5]{Figures/GHGtargetAGR1.png} 
 \caption{Most Ambitious 2030 GHG Targets}
 \label{fig:GHG2030AGR}
\end{figure}

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
A Energy System Optimisation Model (ESOM) can take account of a large number of dynamic variables, and complex interconnections in each time period and each economic sector and determine a least cost solution. The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).  TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
\par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. The expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. \par Previous versions of the Irish TIMES model were extracted from the PET36 model (Pan European TIMES model), and then updated with specific national data and assumptions, these models were used to provide informative reports to Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}. A new TIMES model called ``TIMES Ireland Model" (TIM) was built in 2021, TIM is structured to explore Ireland's new climate targets. 
TIMES has also been widely used to internationally, with some studies focusing on the residential sector. For example, Li studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed how higher thermal efficiency in buildings might hamper co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
TIMES has been used in Ireland and internationally to provide evidence-based and robust decarbonisation insights for climate policy discussions.

@@

The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions. Fig. \ref{fig:Ireland2018GHG}. shows the energy related GHG disaggregated between transport, heat and electricity, where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.

\begin{figure}[!htbp]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
    \hspace{8.5cm}
    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
    \label{fig:Ireland2018GHG}
\end{figure}

 \par The residential sector target is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with other sectors. Despite these high target ambitions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity. The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity and heat, but in transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.

@@

\par

The CAP2019 Marginal Abatement Cost Curve (MACC) developed by McKinsey and Company's modelling tools is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} and provides an analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The CAP2019 MACC in Fig.\ref{fig:CAP2019:MACC} explores technological, behavioural and fuel switching options. The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$. over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. For example, increasing onshore-wind by $8.2 GW$ and offshore-wind would grow to $3.5 GW$,this would be expected to save $7 to 8 MtCO2_e_q$. This would help to indirectly reduce the GHG in the residential sector. Solutions specific to the residential sector are: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{itemize}

Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners. In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology. The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}. The CAP2019 plan to deliver a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.

\subsection{Ireland: Residential Sector Background}
\label{Intro:IreRSDBack}

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, larger dwellings and rural settlement patterns of low thermal efficiency of the housing stock. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, and 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%. \cite{EurostatShareSurveyilc_lvho50a}. The lower population density and large dwellings of the existing Irish housing stock, translates to more expensive GHG mitigation options compared to other residential sectors in the EU.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}. 
In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

@@

%PICK UP HERE
Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock. As mentioned in section \ref{Intro:IreRSD}, Irelands new buildings produce 70\% less GHG emissions than building in 2005, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock highlights the impact the year of construction has on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.The BERs are categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .

@@

\subsection{Current Energy Policy}
\label{Intro:Policy}
Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings
500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent
Installation of 600,000 heat pumps, with 400,000 of these in existing buildings
Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes
Complete the rollout of the Support Scheme for Renewable Heat
Increasing the number of Sustainable Energy Communities to 1,500

@@

\subsection{TIMES Ireland Model (TIM)}
\label{Intro:TIM}
\par
TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets. The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data. The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM . TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing. The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps. TIM will also be open platform to help improve the model and provide transparency. TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy. The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM. for these reasons TIM residential sector is a state of the art modelling platform. 
%GRAPH OF IRELAND DECARBONISATION TARGETS  VS OTHER COUNTRIES - DIFFERENT INTRASTATE and RESOURCE AT BIDENS ENERGY SUBMIT - DENMAKR TOP 2018, 
	ESOM can be used to start the conversation on..
Look at other ESOM, brief explain…
Other residential sectors..
CAP2019 - Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\section{Literature Review}
\label{LitR}
``How should Ireland set GHG targets in the residential sector?” 
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

@@

\section{Sample Appendix Section}
\label{sec:sample:appendix}
XXXXXXXX
